---
title: "DRD_FinalReport_Group1"
author: Amir Aynede, Kian Keshani, Bianca Gustini, Yuri Dobrokhotov, Favour Osese
  Otoijamun, Priscilla Castro Vargas
date: "2025-06-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: Comprehensive Analysis of DNA Methylation Using the Illumina HumanMethylation450 BeadChip Platform
output: html_document
date: "2025-06"
---

# Introduction

# Analytical pipeline and R code

## 0. Preparation (Environment Setting and Importing Libraries)
```{r results="hide", message=FALSE, warning=FALSE}
library(minfi)
library(minfiData)
library(IlluminaHumanMethylation450kmanifest)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(shinyMethyl)
library(AnnotationDbi)
library(sva)
library(qqman)
library(gplots)
install.packages("lib/SummarizedExperiment_1.38.0.tar.gz", repos = NULL)
```

## 1. Load raw data
Set the proper working directory where the data is stored, load raw data with minfi and create an object called RGset storing the RGChannelSet object.

```{r results="hide", message=FALSE, warning=FALSE}
setwd("~/chunche/DRD_2025_Project")
SampleSheet <- read.csv("./data/raw/SampleSheet_Report_II.csv",header=T)
baseDir <- ("./data/raw")
targets <- read.metharray.sheet(baseDir) 
RGset <- read.metharray.exp(targets = targets)
save(RGset,file="RGset_Report.RData")
```

## 2. Creation of R/G dataframes
Create the dataframes Red and Green to store the red and green fluorescences respectively.

**a. Red chanel fluorescence**
Extraction of the fluorescence intensity associated with red.
```{r results="hide", message=FALSE, warning=FALSE}
Red <- data.frame(getRed(RGset)) 
dim(Red)
head(Red)
```

**b. Green chanel fluorescence**
Extraction of the fluorescence intensity associated with green.
```{r results="hide", message=FALSE, warning=FALSE}
Green <- data.frame(getGreen(RGset))
dim(Green)
head(Green)
```

## 3. Examination of  a probe 
Determine whether address 18756452 belongs to a Type I or Type II probe by retrieving the red and green fluorescence values for this address and checking the manifest file.

Creation of data frames containing the result of Type I and Type II probes respectively.
```{r results="hide", message=FALSE, warning=FALSE}
ProbeInfo_I <- data.frame(getProbeInfo(RGset))
ProbeInfo_II <- data.frame(getProbeInfo(RGset, type = "II"))
```

Check the probe having the address 18756452 in the Type I array:
```{r results="hide", message=FALSE, warning=FALSE}
ProbeInfo_I[ProbeInfo_I$AddressA=="18756452",]
ProbeInfo_I[ProbeInfo_I$AddressB=="18756452",]
```

Check the probe having the address 18756452 in the Type I array:
```{r results="hide", message=FALSE, warning=FALSE}
ProbeInfo_II[ProbeInfo_II$AddressB=="18756452",] 
ProbeInfo_II[ProbeInfo_II$AddressA=="18756452",]
```

++ FILL THE FOLLOWING TABLE.

## 4. Creation of the object MSet.raw
```{r results="hide", message=FALSE, warning=FALSE}
MSet.raw <- preprocessRaw(RGset)
save(MSet.raw,file="./input_data/MSet_raw.RData")
Meth <- as.matrix(getMeth(MSet.raw))
Unmeth <- as.matrix(getUnmeth(MSet.raw))
```

## 5. Quality control
Perform quality checks and provide a brief comment to each step. 

### a. QCplot
```{r message=FALSE, warning=FALSE}
pdf("QCplot.pdf")
qc <- getQC(MSet.raw)
plotQC(qc)
dev.off()
```
++COMMENT.

### b. Intensity of negative controls
```{r message=FALSE, warning=FALSE}
getProbeInfo(RGset, type = "Control")
NegativeControl <- data.frame(getProbeInfo(RGset, type = "Control"))
table(NegativeControl$Type)
```
++COMMENT.

### c. Calculation of detection p-values
```{r message=FALSE, warning=FALSE}
Detection_pvalue <- detectionP(RGset)
save(Detection_pvalue,file="Detection_pvalue.RData")
Over_Threshold <- Detection_pvalue>0.05
head(Over_Threshold)
table(Over_Threshold)
sum(Over_Threshold)
paste0(sprintf('There are %s failed probes with a detection p-value higher than 0.05', sum(failed)))
Over_Threshold_Per_Sample <- colSums(Over_Threshold)
Over_Threshold_Per_Sample
summary(Over_Threshold)
```
++COMMENT.
++how many probes have a detection p-value higher than 0.05?


## 6. Calculation of raw Beta and M values
Calculate raw beta and M values and plot the densities of mean methylation values, dividing the samples in CTRL and DIS (suggestion: subset the beta and M values matrixes in order to retain CTRL or DIS subjects and apply the function mean to the 2 subsets). 

```{r message=FALSE, warning=FALSE}
beta <- getBeta(MSet.raw)
M <- getM(MSet.raw)
beta_df <- data.frame(beta)
M_df <- data.frame(M)

pheno <-SampleSheet
pheno$Group
beta_ctrl <- beta_df[pheno$Group=="CTRL",]
beta_dis <- beta_df[pheno$Group=="DIS",]
M_ctrl <- M_df[pheno$Group=="CTRL",]
M_dis <- M_df[pheno$Group=="DIS",]


mean_of_beta_ctrl <- apply(beta_ctrl,1,mean,na.rm=T)
mean_of_beta_dis <- apply(beta_dis,1,mean,na.rm=T)
mean_of_M_ctrl <- apply(M_ctrl,1,mean,na.rm=T)
mean_of_M_dis <- apply(M_dis,1,mean,na.rm=T)
```

Plotting the density of mean methylation values.
```{r message=FALSE, warning=FALSE}
pdf("Density_Mean_Methylation_Values.pdf")
par(mfrow=c(1,2))
plot(density(mean_of_beta_ctrl,na.rm=T), col="orange", lwd=2, main="Density of Mean Beta Values")
plot(density(mean_of_M_ctrl, na.rm=T), col="purple")
plot(density(mean_of_beta_dis,na.rm=T), col="orange")
plot(density(mean_of_M_dis, na.rm=T), col="purple")
dev.off()
```

++ Do you see any difference between the two groups?

## 7. Data normalization
Normalize the data using preprocessNoob and compare raw data and normalized data. Produce a plot with 6 panels in which, for both raw and normalized data, you show the density plots of beta mean values according to the chemistry of the probes, the density plot of beta standard deviation values according to the chemistry of the probes and the boxplot of beta values. 

Subset the data frame of beta values according to Type I and Type II.
```{r message=FALSE, warning=FALSE}
dfI <- Illumina450Manifest_clean[Illumina450Manifest_clean$Infinium_Design_Type=="I",]
dfI <- droplevels(dfI)
dfII <- Illumina450Manifest_clean[Illumina450Manifest_clean$Infinium_Design_Type=="II",]
dfII <- droplevels(dfII)

beta_I <- beta[rownames(beta) %in% dfI$IlmnID,]
beta_II <- beta[rownames(beta) %in% dfII$IlmnID,]

mean_of_beta_I <- apply(beta_I,1,mean)
mean_of_beta_II <- apply(beta_II,1,mean)
d_mean_of_beta_I <- density(mean_of_beta_I,na.rm=T)
d_mean_of_beta_II <- density(mean_of_beta_II,na.rm=T)
```

Calculation of the densities of the standard deviations, which can be calculated using the function sd():
```{r results="hide", message=FALSE, warning=FALSE}
sd_of_beta_I <- apply(beta_I,1,sd,na.rm=T)
sd_of_beta_II <- apply(beta_II,1,sd,na.rm=T)
d_sd_of_beta_I <- density(sd_of_beta_I,)
d_sd_of_beta_II <- density(sd_of_beta_II) #doesn't work. We need to find out why- all na values should have been omited
d_sd_of_beta_II <- density(na.omit(sd_of_beta_II))
```

Use of preprocessNoob.
```{r results="hide", message=FALSE, warning=FALSE}
load("~/Downloads/L3/RGset.RData")
preprocessNoob_results <- preprocessNoob(RGset)
beta_preprocessNoob <- getBeta(preprocessNoob_results)
save(beta_preprocessNoob, file="beta_preprocessNoob.RData")
```

Divide beta_preprocessNoob matrix according to type I and type II probes, calculate the mean and the standartd deviation for each probe across the 8 samples and calculate the density distributions.
```{r results="hide", message=FALSE, warning=FALSE}
beta_preprocessNoob_I <- beta_preprocessNoob[rownames(beta_preprocessNoob) %in% dfI$IlmnID,]
beta_preprocessNoob_II <- beta_preprocessNoob[rownames(beta_preprocessNoob) %in% dfII$IlmnID,]
mean_of_beta_preprocessNoob_I <- apply(beta_preprocessNoob_I,1,mean)
mean_of_beta_preprocessNoob_II <- apply(beta_preprocessNoob_II,1,mean)
d_mean_of_beta_preprocessNoob_I <- density(mean_of_beta_preprocessNoob_I,na.rm=T)
d_mean_of_beta_preprocessNoob_II <- density(mean_of_beta_preprocessNoob_II,na.rm=T)
sd_of_beta_preprocessNoob_I <- apply(beta_preprocessNoob_I,1,sd)
sd_of_beta_preprocessNoob_II <- apply(beta_preprocessNoob_II,1,sd)
d_sd_of_beta_preprocessNoob_I <- density(sd_of_beta_preprocessNoob_I,na.rm=T)
d_sd_of_beta_preprocessNoob_II <- density(sd_of_beta_preprocessNoob_II,na.rm=T)
```

Plot the data.
```{r results="hide", message=FALSE, warning=FALSE}
pdf("Plot_comparison_raw_preprocessNoob.pdf",height=7,width=15)
par(mfrow=c(2,3))
plot(d_mean_of_beta_I, col="blue", main="Raw Beta", ylim=c(0, 5))
lines(d_mean_of_beta_II, col="red")
plot(d_sd_of_beta_I, col="blue", main="Raw SD", ylim=c(0, 30))
lines(d_sd_of_beta_II, col="red")
boxplot(beta, main="Raw Beta Boxplot", ylim=c(0, 1))

plot(d_mean_of_beta_preprocessNoob_I, col="blue", main="PreprocessNoob Beta", ylim=c(0, 5))
lines(d_mean_of_beta_preprocessNoob_II, col="red")
plot(d_sd_of_beta_preprocessNoob_I, col="blue", main="PreprocessNoob SD", ylim=c(0, 60))
lines(d_sd_of_beta_preprocessNoob_II, col="red")
boxplot(beta_preprocessNoob, main="PreprocessNoob Beta Boxplot", ylim=c(0, 1))
```

Boxplot of normalized values.
```{r results="hide", message=FALSE, warning=FALSE}
pheno$Group <- as.factor(pheno$Group)
```

Boxplot of normalized values.
```{r results="hide", message=FALSE, warning=FALSE}
par(mfrow=c(1,1))
boxplot(beta_preprocessNoob,
        las=2,                      # rotate x-axis labels
        col=c("skyblue", "green")[pheno$Group],  # color by group
        ylab='Beta values',
        xlab='Samples',
        main='Boxplot of Normalised Beta Values')

legend("topright", legend=levels(pheno$Group), fill=c("skyblue", "green"))
```


Plot based on sample.

```{r results="hide", message=FALSE, warning=FALSE}
pdf("Plot_comparison_raw_preprocessNoob control vs disease.pdf",height=7,width=15)
par(mfrow=c(2,3))
plot(d_mean_of_beta_I,col="blue",main="raw beta")
lines(d_mean_of_beta_II,col="red")
plot(d_sd_of_beta_I,col="blue",main="raw sd")
lines(d_sd_of_beta_II,col="red")
boxplot(beta,
        las=2,                      # rotate x-axis labels
        col=c("skyblue", "green")[pheno$Group],  # color by group
        ylab='Beta values',
        xlab='Samples',
        main='Boxplot of Beta Values')

plot(d_mean_of_beta_preprocessNoob_I,col="blue",main="preprocessNoob beta")
lines(d_mean_of_beta_preprocessNoob_II,col="red")
plot(d_sd_of_beta_preprocessNoob_I,col="blue",main="preprocessNoob sd")
lines(d_sd_of_beta_preprocessNoob_II,col="red")
boxplot(beta_preprocessNoob,
        las=2,                      # rotate x-axis labels
        col=c("skyblue", "green")[pheno$Group],  # color by group
        ylab='Beta values',
        xlab='Samples',
        main='Boxplot of Normalised Beta Values')
dev.off()
```

++Provide a short comment about the changes you observe.
++Do you think that the normalization approach that you used is appropriate considering this specific dataset? Try to color the boxplots according to the group (WT and MUT) and check whether the distribution of methylation values is different between the two groups, before and after normalization … 

## 8. Principal Component Analysis
Perform a PCA on the matrix of normalized beta values generated in step 7. 

```{r results="hide", message=FALSE, warning=FALSE}
pca_results <- prcomp(t(beta_preprocessNoob),scale=T)
par(mfrow=c(1,1))
plot(pca_results,main="PCA results",ylim=c(0,150000))
pca_results$x

plot(pca_results$x[,1],pca_results$x[,2],cex=1,pch=2,xlab="PC1",ylab="PC2",main='PCA', xlim=c(-1000,1000),ylim=c(-1000,1000))
text(pca_results$x[,1], pca_results$x[,2], labels=(pheno$SampleID), cex=0.7, pos=1)

pheno$Group
palette(c('orange','green'))
plot(pca_results$x[,1],pca_results$x[,2],cex=1,pch=17,col=pheno$Group,main='PCA by group',xlab="PC1",ylab="PC2",xlim=c(-1000,1000),ylim=c(-1000,1000))
text(pca_results$x[,1], pca_results$x[,2], labels=(pheno$SampleID), cex=0.7, pos=1)
legend('bottomright', legend=levels(pheno$Group), col=c(1:nlevels(pheno$Group)), pch=17)

pheno$Sex
pheno$Sex <- as.factor(pheno$Sex)
palette(c('pink','aquamarine'))
plot(pca_results$x[,1],pca_results$x[,2],cex=2,pch=17,col=pheno$Sex,main='PCA by sex',xlab="PC1",ylab="PC2",xlim=c(-1000,1000),ylim=c(-1000,1000))
text(pca_results$x[,1], pca_results$x[,2], labels=(pheno$SampleID), cex=0.7, pos=1)
legend('bottomright', legend=levels(pheno$Sex), col=c(1:nlevels(pheno$Sex)), pch=17)

pheno$Sentrix_ID
pheno$Sentrix_ID <- as.factor(pheno$Sentrix_ID)
palette(c('red','blue','green','yellow','purple'))
plot(pca_results$x[,1],pca_results$x[,2],cex=2,pch=17,col=pheno$Sentrix_ID,main='PCA by Sentrix_ID',xlab="PC1",ylab="PC2",xlim=c(-1000,1000),ylim=c(-1000,1000))
text(pca_results$x[,1], pca_results$x[,2], labels=(pheno$SampleID), cex=0.7, pos=1)
legend('bottomright', legend=levels(pheno$Sentrix_ID), col=c(1:nlevels(pheno$Sentrix_ID)), pch=17)
```

++Comment the plot.
++Do you see any outlier? Do the samples divide according to the group? Do they divide according to the sex of the samples? Do they divide according to the batch, that is the column Sentrix_ID?). 

## 9. Identification of differentially methylated probes
Using the matrix of normalized beta values generated in step 7, identify differentially methylated probes between CTRL and DIS groups using a t-test.

```{r results="hide", message=FALSE, warning=FALSE}
t_test <- t.test(beta_preprocessNoob[1,] ~ pheno$Group)
```

We are interested in the p.value element of these lists. We use the apply() function to calculate the mean of each row of the matrix. However, unlike the mean() function, t.test() and wilcox.test() functions do not return a value, but a list; therefore, we have to create an ad hoc function:
```{r results="hide", message=FALSE, warning=FALSE}
My_ttest_function <- function(x) {
  t_test <- t.test(x~ pheno$Group)
  return(t_test$p.value)
} 
```

Use of the function.
```{r results="hide", message=FALSE, warning=FALSE}
first20k_beta_preprocessNoob <- beta_preprocessNoob[1:20000,]
pValues_ttest_20k <- apply(first20k_beta_preprocessNoob,1, My_ttest_function)
pValues_ttest_20k
```

Creation of a data frame with all the beta values and the pValue column.
```{r results="hide", message=FALSE, warning=FALSE}
final_ttest_20k <- data.frame(first20k_beta_preprocessNoob, pValues_ttest_20k)
```

Organization of the probes on the basis of the pValue column.
```{r results="hide", message=FALSE, warning=FALSE}
final_ttest_20k <- final_ttest_20k[order(final_ttest_20k$pValues_ttest_20k),]
head(final_ttest_20k)
```

Filter for significant probes
```{r results="hide", message=FALSE, warning=FALSE}
significant_ttest <- final_ttest_20k[final_ttest_20k$pValues_ttest_20k <= 0.05, ]
```

## 10. Multiple test correction
Apply multiple test correction and set a significant threshold of 0.05.

```{r results="hide", message=FALSE, warning=FALSE}
corrected_pValues_BH <- p.adjust(final_ttest_20k$pValues_ttest_20k,"BH")
corrected_pValues_Bonf <- p.adjust(final_ttest_20k$pValues_ttest_20k,"bonferroni")
final_ttest_20k_corrected <- data.frame(final_ttest_20k, corrected_pValues_BH, corrected_pValues_Bonf)
head(finalttest_20k_corrected)
```

Visualization of the distributions of the p-values by boxplots.
```{r results="hide", message=FALSE, warning=FALSE}
colnames(final_ttest_20k_corrected)
boxplot(final_ttest_20k_corrected[,9:11])
```

Determining the amount of probes that remain after the MTC.
```{r results="hide", message=FALSE, warning=FALSE}
dim(final_ttest_20k_corrected[final_ttest_20k_corrected$pValues_ttest<=0.05,])
dim(final_ttest_20k_corrected[final_ttest_20k_corrected$corrected_pValues_BH<=0.05,])
dim(final_ttest_20k_corrected[final_ttest_20k_corrected$corrected_pValues_Bonf<=0.05,])
```

++How many probes do you identify as differentially methylated considering nominal pValues? How many after Bonferroni correction? How many after BH correction?

## 11. Graphic the results of the differential methylation analysis.
Produce a volcano plot and a Manhattan plot of the results of differential methylation analysis. 

Preparation of the data.
```{r results="hide", message=FALSE, warning=FALSE}
beta_first20k <- final_ttest_20k_corrected[,1:8]
beta_first20k_ctrl <- beta_first20k[,pheno$Group=="CTRL"]
mean_beta_first20k_ctrl <- apply(beta_first20k_ctrl,1,mean)
beta_first20k_dis <- beta_first20k[,pheno$Group=="DIS"]
mean_beta_first20k_dis <- apply(beta_first20k_dis,1,mean)
```

Calculation of the difference between average values.
```{r results="hide", message=FALSE, warning=FALSE}
delta_first20k <- mean_beta_first20k_dis-mean_beta_first20k_ctrl
head(delta_first20k)
```

Creation of a data frame.
```{r results="hide", message=FALSE, warning=FALSE}
toVolcPlot <- data.frame(delta_first20k, -log10(final_ttest_20k_corrected$pValues_ttest))
head(toVolcPlot)
plot(toVolcPlot[,1], toVolcPlot[,2])
```

**a. Volcano plot**
Highlight the probes with a nominal pValue<0.01 and a delta > 0.1.
```{r results="hide", message=FALSE, warning=FALSE}
pdf("Volcano_plot.pdf")
plot(toVolcPlot[,1], toVolcPlot[,2],pch=16,cex=0.5,main='volcano plot')
toHighlight <- toVolcPlot[toVolcPlot[,1]>0.1 & toVolcPlot[,2]>(-log10(0.05)),]
head(toHighlight)
points(toHighlight[,1], toHighlight[,2],pch=16,cex=0.7,col="red")
dev.off()
```

**b. Manhattan plot**

Load of the result of a reduced t-test and the Manifest and merge this objects.
```{r results="hide", message=FALSE, warning=FALSE}
load("~/Downloads/L6/final_ttest_reduced.RData")
load('~/Downloads/L2/Illumina450Manifest_clean.RData')
final_ttest_reduced <- data.frame(rownames(final_ttest_reduced),final_ttest_reduced)
head(final_ttest_reduced)
colnames(final_ttest_reduced)
colnames(final_ttest_reduced)[1] <- "IlmnID"
colnames(final_ttest_reduced)

final_ttest_reduced_annotated <- merge(final_ttest_reduced, Illumina450Manifest_clean,by="IlmnID")
dim(final_ttest_reduced)
dim(Illumina450Manifest_clean)
dim(final_ttest_reduced_annotated)
str(final_ttest_reduced_annotated)
```

Creation of the input fot the Manhattan plot.
```{r results="hide", message=FALSE, warning=FALSE}
input_Manhattan <- final_ttest_reduced_annotated[colnames(final_ttest_reduced_annotated) %in% c("IlmnID","CHR","MAPINFO","pValues_ttest")]
dim(input_Manhattan)
head(input_Manhattan)
str(input_Manhattan$CHR)
levels(input_Manhattan$CHR)
order_chr <- c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15","16","17","18","19","20","21","22","X","Y")
input_Manhattan$CHR <- factor(input_Manhattan$CHR,levels=order_chr )
levels(input_Manhattan$CHR)

```

Creation of the Manhattan plot.
```{r results="hide", message=FALSE, warning=FALSE}
input_Manhattan$CHR <- as.numeric(input_Manhattan$CHR)
table(input_Manhattan$CHR)
manhattan(input_Manhattan, snp="IlmnID",chr="CHR", bp="MAPINFO", p="pValues_ttest" )
-log10(0.00001)
pdf("Manhattan_plot.pdf")
manhattan(input_Manhattan, snp="IlmnID",chr="CHR", bp="MAPINFO", p="pValues_ttest",annotatePval = 0.00001,col=rainbow(24) )
dev.off()
```


## 12. Heatmap
Produce an heatmap of the top 100 significant, differentially methylated probes.
```{r results="hide", message=FALSE, warning=FALSE}
pheno$Group
colorbar <- c("green","orange","orange","green","orange","orange","green","green")
```

Compared the results of hierchical clustering using different methods.
```{r results="hide", message=FALSE, warning=FALSE}
input_heatmap=as.matrix(final_ttest_20k[1:100,1:8])
heatmap.2(input_heatmap,col=terrain.colors(100),Rowv=T,Colv=T,dendrogram="both",key=T,ColSideColors=colorbar,density.info="none",trace="none",scale="none",symm=F,main="Complete linkage")
```

Single
```{r results="hide", message=FALSE, warning=FALSE}
heatmap.2(input_heatmap,col=terrain.colors(100),Rowv=T,Colv=T,hclustfun = function(x) hclust(x,method = 'single'),dendrogram="both",key=T,ColSideColors=colorbar,density.info="none",trace="none",scale="none",symm=F,main="Single linkage")
```

Average
```{r results="hide", message=FALSE, warning=FALSE}
heatmap.2(input_heatmap,col=terrain.colors(100),Rowv=T,Colv=T,hclustfun = function(x) hclust(x,method = 'average'),dendrogram="both",key=T,ColSideColors=colorbar,density.info="none",trace="none",scale="none",symm=F,main="Average linkage")
```

Adjustment of the colors.
```{r results="hide", message=FALSE, warning=FALSE}
col2=colorRampPalette(c("green","black","red"))(100)
heatmap.2(input_heatmap,col=col2,Rowv=T,Colv=T,dendrogram="both",key=T,ColSideColors=colorbar,density.info="none",trace="none",scale="none",symm=F)
```
